---
title: "Assignment 5: Data Visualization"
author: "Damon L. Daniels"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 

library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

here()

nutrients <- read.csv("./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
litter <- read.csv("./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 

litter$collectDate <- ymd(litter$collectDate)
nutrients$sampledate <- ymd(nutrients$sampledate)


```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

DLDtheme <- theme_gray(base_size = 9) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7),
        panel.grid = element_line(color = "blue",
                                  size = 0.5)
        )

theme_set(DLDtheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

tp_by_po4 <-
  ggplot(nutrients, aes(x = tp_ug, y = po4, color = lakename)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(color = "Lake Name") +
  xlim(0, 150) +
  ylim(0, 50) +
  DLDtheme
print(tp_by_po4)

# or,

tp_by_po4 <-
  ggplot(nutrients, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(color = "Lake Name") +
  xlim(0, 50) +
  ylim(0, 150) +
  DLDtheme
print(tp_by_po4)

# this is the correct plot (y by x)


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

DLDtheme2 <- theme_gray(base_size = 9) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text())
        legend.position = "top"
        legend.text = element_text(size = 7)
        legend.title = element_text(size = 7)
        panel.grid = element_line(color = "blue",
                                  size = 0.5)
theme_set(DLDtheme2)

nutrients$month_factor <- factor(
  nutrients$month,
  levels = 1:12,
  labels = month.abb) 

temp_by_month <-
  ggplot(nutrients, mapping = aes(
    x = month_factor, y = temperature_C, color = lakename)) +
  theme(legend.position = "none") +
  xlab("Month") +
  ylab(expression(paste("Temperature (", degree, "C)"))) +
  geom_boxplot() +
  labs(color = "Lake Name") +
  ylim(0, 30)
print(temp_by_month)

tp_by_month <-
  ggplot(nutrients, mapping = aes(
    x = month_factor, y = tp_ug, color = lakename)) +
  theme(legend.position = "none") +
  xlab("Month") +
  ylab("TP") +
  geom_boxplot() +
  labs(color = "Lake Name") +
  ylim(0, 150)
print(tp_by_month)

tn_by_month <-
  ggplot(nutrients, mapping = aes(
    x = month_factor, y = tn_ug, color = lakename)) +
  theme(legend.position = "bottom") +
  xlab("Month") +
  ylab("TN") +
  geom_boxplot() +
  labs(color = "Lake Name  ") +
  ylim(0, 3000)
print(tn_by_month)

plot_grid(temp_by_month, tp_by_month, tn_by_month, nrow = 3, rel_heights = c(1, 1, 1))
aligned <- align_plots(temp_by_month, tp_by_month, tn_by_month, align = "vh")
plot_grid(temp_by_month, tp_by_month, tn_by_month, ncol = 1, align = "vh")



```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: For both Paul Lake and Peter Lake: temperatures, and both TP and TN 
levels, trend upward as summer approaches, before peaking in September and 
dropping pretty significantly around the beginning of fall. For temperature,
this drop continues across the succeeding month. When comparing between
Paul Lake and Peter Lake: temperatures trend higher for Paul Lake from May
through August; temperatures are approximately even between the two lakes in September, and Peter Lake has higher temperatures in October and November.
Both TP and TN levels are higher at Peter Lake for each of the months observed,
which are the late spring and full summer months.  

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

litter_needles_subset <-
  ggplot(subset(litter, functionalGroup == "Needles"),
         aes(x = collectDate, y = dryMass, color = nlcdClass)
         ) +
  geom_point()
print(litter_needles_subset)

#7

litter_needles_subset <-
  ggplot(subset(litter, functionalGroup == "Needles"),
         aes(x = collectDate, y = dryMass)
         ) +
  facet_wrap(vars(nlcdClass), nrow = 3, ncol = 1) +
  geom_point()
print(litter_needles_subset)


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think plot 7 is is the more effective plot. 
The separation by nlcdClass allows more observations to be determined,
in my view. One can compare across the classes of Needles to get an easier
picture of (1) which accumulates more dryMass, and tacitly, during which relative times within the years that are graphed.