---
title: "Assignment 2: Coding Basics"
author: Damon L. Daniels
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

```{r}
sequence_1_to_30_by_3s <- seq(1,30,3) #sequence from 1 to 30, by 3s
print(sequence_1_to_30_by_3s)
```

2.  Compute the mean and median of this sequence.

```{r}
mean (sequence_1_to_30_by_3s) # find mean of sequence = 14.5
median (sequence_1_to_30_by_3s) # find median of sequence = 14.5

# We are writing codes that give us sequences, assigning new names for codes, 
# determining means and medians, and determining conditional statements.

```

3.  Ask R to determine whether the mean is greater than the median.

```{r}

mean(sequence_1_to_30_by_3s) != median(sequence_1_to_30_by_3s) # are the mean 
# and median equal (TRUE or FALSE)?
mean(sequence_1_to_30_by_3s) == median(sequence_1_to_30_by_3s) # are the mean 
# and median equal (TRUE or FALSE)?
mean(sequence_1_to_30_by_3s) < median(sequence_1_to_30_by_3s) # is the mean less 
# than the median (TRUE or FALSE)?
mean(sequence_1_to_30_by_3s) > median(sequence_1_to_30_by_3s) # is the mean 
# greater than the median (TRUE or FALSE)?

```

4.  Insert comments in your code to describe what you are doing.

```{r}
# We are writing codes that give us sequences, assigning new names for codes,
# determining means and medians, and determining conditional statements to compare the mean and median values.
```

```{r basics 1}
#1. 

sequence_1_to_30_by_3s <- seq(1,30,3)

#2. 

mean (sequence_1_to_30_by_3s)
median (sequence_1_to_30_by_3s)

#3. 

mean(sequence_1_to_30_by_3s) != median(sequence_1_to_30_by_3s)
mean(sequence_1_to_30_by_3s) == median(sequence_1_to_30_by_3s) 
mean(sequence_1_to_30_by_3s) < median(sequence_1_to_30_by_3s) 
mean(sequence_1_to_30_by_3s) > median(sequence_1_to_30_by_3s) 

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

```{r}

test_scores <- c(75, 40, 97, 86, 34) #numeric operator (test scores)
student_name <- c("Damon Daniels",
                  "Cornell Jordan",
                  "Aarushi Tripathi",
                  "Maria Mohyuddin",
                  "Angeline Yin-Chia") #character operator (name of student)
test_passed <- test_scores >= 50 #logical operator (test score threshold)

```

6.  Label each vector with a comment on what type of vector it is.

```{r}
# Please see above.
```

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

```{r}

df_test_outcome <- data.frame(
  "NAME" = student_name,
  "SCORE" = test_scores,
  "STUDENT_PASSED" = test_passed
)

print(df_test_outcome)

```

8.  Label the columns of your data frame with informative titles.

```{r basics 2}

# Please see above.

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame contains multiple types/categories of data, whereas a matrix can only contain one. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

```{r basics 3}

PASSorFAIL <- function(x){
  thegrade <- ifelse(x>=50,TRUE,FALSE)
  print(thegrade)
}

```



11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 4}
PASSorFAIL(df_test_outcome$SCORE)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: "if" and "else" did not work b/c there were more than one test score.

